---
title: "Movie Recommendation System 1813033_1813042"
output:
  html_document: default
  pdf_document: default
---
Problem statement: 
The main goal of this machine learning project is to build a recommendation engine that recommends movies to users.

Recommendation System:
A recommendation system provides suggestions to the users through a filtering process that is based on user preferences and browsing history. The information about the user is taken as an input. The information is taken from the input that is in the form of browsing data. This information reflects the prior usage of the product as well as the assigned ratings. A recommendation system is a platform that provides its users with various contents based on their preferences and likings. The recommendation system is an implementation of the machine learning algorithms.

Method Used:
Collaborative filtering:
Collaborative filtering approaches build a model from user’s past behavior (i.e. items purchased or searched by the user) as well as similar decisions made by other users. This model is then used to predict items (or ratings for items) that user may have an interest in.

Technique Used: Regression

```{r setup, include=FALSE}

library(recommenderlab)
library(ggplot2)                       
library(data.table)
library(reshape2)
```

```{r}
setwd("C:/Users/FG/Desktop/movies")                       
movie_data <- read.csv("movies.csv",stringsAsFactors=FALSE)
rating_data <- read.csv("ratings.csv")
str(movie_data)



```


```{r movies}
summary(movie_data)
head(movie_data)
```

```{r}
summary(rating_data)
head(rating_data)
```
data preprocessing
```{r}
movie_genre <- as.data.frame(movie_data$genres, stringsAsFactors=FALSE)
library(data.table)
movie_genre2 <- as.data.frame(tstrsplit(movie_genre[,1], '[|]', 
                                   type.convert=TRUE), 
                         stringsAsFactors=FALSE) #DataFlair
colnames(movie_genre2) <- c(1:10)

list_genre <- c("Action", "Adventure", "Animation", "Children", 
                "Comedy", "Crime","Documentary", "Drama", "Fantasy",
                "Film-Noir", "Horror", "Musical", "Mystery","Romance",
                "Sci-Fi", "Thriller", "War", "Western")
genre_mat1 <- matrix(0,10330,18)
genre_mat1[1,] <- list_genre
colnames(genre_mat1) <- list_genre

for (index in 1:nrow(movie_genre2)) {
  for (col in 1:ncol(movie_genre2)) {
    gen_col = which(genre_mat1[1,] == movie_genre2[index,col]) #Author DataFlair
    genre_mat1[index+1,gen_col] <- 1
}
}
genre_mat2 <- as.data.frame(genre_mat1[-1,], stringsAsFactors=FALSE) #remove first row, which was the genre list
for (col in 1:ncol(genre_mat2)) {
  genre_mat2[,col] <- as.integer(genre_mat2[,col]) #convert from characters to integers
} 
str(genre_mat2)

```

In the next step of Data Pre-processing of R project, we will create a ‘search matrix’ that will allow us to perform an easy search of the films by specifying the genre present in our list.

```{r}
SearchMatrix <- cbind(movie_data[,1:2], genre_mat2[])
head(SearchMatrix)    
```

There are movies that have several genres, for example, Toy Story, which is an animated film also falls under the genres of Comedy, Fantasy, and Children. This applies to the majority of the films.

For our movie recommendation system to make sense of our ratings through recommenderlabs, we have to convert our matrix into a sparse matrix one. This new matrix is of the class ‘realRatingMatrix’. This is performed as follows:

```{r}
ratingMatrix <- dcast(rating_data, userId~movieId, value.var = "rating", na.rm=FALSE)
ratingMatrix <- as.matrix(ratingMatrix[,-1]) #remove userIds
#Convert rating matrix into a recommenderlab sparse matrix
ratingMatrix <- as(ratingMatrix, "realRatingMatrix")
ratingMatrix
```

Let us now overview some of the important parameters that provide us various options for building recommendation systems for movies-

```{r}
recommendation_model <- recommenderRegistry$get_entries(dataType = "realRatingMatrix")
names(recommendation_model)
```

```{r}
lapply(recommendation_model, "[[", "description")
```



```{r}
recommendation_model$IBCF_realRatingMatrix$parameters
```

Exploring Similar Data
Collaborative Filtering involves suggesting movies to the users that are based on collecting preferences from many other users. For example, if a user A likes to watch action films and so does user B, then the movies that the user B will watch in the future will be recommended to A and vice-versa. Therefore, recommending movies is dependent on creating a relationship of similarity between the two users. With the help of recommenderlab, we can compute similarities using various operators like cosine, pearson as well as jaccard.


```{r}
similarity_mat <- similarity(ratingMatrix[1:4, ],
                               method = "cosine",
                               which = "users")
as.matrix(similarity_mat)

image(as.matrix(similarity_mat), main = "User's Similarities")
```


In the above matrix, each row and column represents a user. We have taken four users and each cell in this matrix represents the similarity that is shared between the two users.

Now, we delineate the similarity that is shared between the films –

```{r}
movie_similarity <- similarity(ratingMatrix[, 1:4], method =
                                 "cosine", which = "items")
as.matrix(movie_similarity)

image(as.matrix(movie_similarity), main = "Movies similarity")
```



Let us now extract the most unique ratings –
```{r}
rating_values <- as.vector(ratingMatrix@data)
unique(rating_values) # extracting unique ratings
```

Now, we will create a table of ratings that will display the most unique ratings.

```{r}
Table_of_Ratings <- table(rating_values) # creating a count of movie ratings
Table_of_Ratings
```

Most Viewed Movies Visualization
In this section of the machine learning project, we will explore the most viewed movies in our dataset. We will first count the number of views in a film and then organize them in a table that would group them in descending order.

```{r}
library(ggplot2)
movie_views <- colCounts(ratingMatrix) # count views for each movie
table_views <- data.frame(movie = names(movie_views),
                          views = movie_views) # create dataframe of views
table_views <- table_views[order(table_views$views,
                                 decreasing = TRUE), ] # sort by number of views
table_views$title <- NA
for (index in 1:10325){
  table_views[index,3] <- as.character(subset(movie_data,
                                         movie_data$movieId == table_views[index,1])$title)
}
table_views[1:6,]
```

Now, we will visualize a bar plot for the total number of views of the top films. We will carry this out using ggplot2.
```{r}
ggplot(table_views[1:6, ], aes(x = title, y = views)) +
  geom_bar(stat="identity", fill = 'steelblue') +
  geom_text(aes(label=views), vjust=-0.3, size=3.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +

  ggtitle("Total Views of the Top Films")
```


Heatmap of Movie Ratings
Now, in this data science project of Recommendation system, we will visualize a heatmap of the movie ratings. This heatmap will contain first 25 rows and 25 columns as follows –

```{r}
image(ratingMatrix[1:20, 1:25], axes = FALSE, main = "Heatmap of the first 25 rows and 25 columns")
```

Performing Data Preparation
We will conduct data preparation in the following three steps –

Selecting useful data.
Normalizing data.
Binarizing the data.
For finding useful data in our dataset, we have set the threshold for the minimum number of users who have rated a film as 50. This is also same for minimum number of views that are per film. This way, we have filtered a list of watched films from least-watched ones.

```{r}
movie_ratings <- ratingMatrix[rowCounts(ratingMatrix) > 50,
                             colCounts(ratingMatrix) > 50]
movie_ratings
```

From the above output of ‘movie_ratings’, we observe that there are 420 users and 447 films as opposed to the previous 668 users and 10325 films. We can now delineate our matrix of relevant users as follows 

```{r}
minimum_movies<- quantile(rowCounts(movie_ratings), 0.98)
minimum_users <- quantile(colCounts(movie_ratings), 0.98)
image(movie_ratings[rowCounts(movie_ratings) > minimum_movies,
                     colCounts(movie_ratings) > minimum_users],
main = "Heatmap of the top users and movies")
```

Now, we will visualize the distribution of the average ratings per user.

```{r}
average_ratings <- rowMeans(movie_ratings)
qplot(average_ratings, fill=I("steelblue"), col=I("red")) +
  ggtitle("Distribution of the average rating per user")

```

Data Normalization
In the case of some users, there can be high ratings or low ratings provided to all of the watched films. This will act as a bias while implementing our model. In order to remove this, we normalize our data. Normalization is a data preparation procedure to standardize the numerical values in a column to a common scale value. This is done in such a way that there is no distortion in the range of values. Normalization transforms the average value of our ratings column to 0. We then plot a heatmap that delineates our normalized ratings.

```{r}
normalized_ratings <- normalize(movie_ratings)
sum(rowMeans(normalized_ratings) > 0.00001)

image(normalized_ratings[rowCounts(normalized_ratings) > minimum_movies,
                          colCounts(normalized_ratings) > minimum_users],
main = "Normalized Ratings of the Top Users")
```



Performing Data Binarization
In the final step of our data preparation in this data science project, we will binarize our data. Binarizing the data means that we have two discrete values 1 and 0, which will allow our recommendation systems to work more efficiently. We will define a matrix that will consist of 1 if the rating is above 3 and otherwise it will be 0.

```{r}
binary_minimum_movies <- quantile(rowCounts(movie_ratings), 0.95)
binary_minimum_users <- quantile(colCounts(movie_ratings), 0.95)
#movies_watched <- binarize(movie_ratings, minRating = 1)

good_rated_films <- binarize(movie_ratings, minRating = 3)
image(good_rated_films[rowCounts(movie_ratings) > binary_minimum_movies,
colCounts(movie_ratings) > binary_minimum_users],
main ="Heatmap of the top users and movies")
```

Collaborative Filtering System
In this section of data science project, we will develop our very own Item Based Collaborative Filtering System. This type of collaborative filtering finds similarity in the items based on the people’s ratings of them. The algorithm first builds a similar-items table of the customers who have purchased them into a combination of similar items. This is then fed into the recommendation system.

The similarity between single products and related products can be determined with the following algorithm –

For each Item i1 present in the product catalog, purchased by customer C.
And, for each item i2 also purchased by the customer C.
Create record that the customer purchased items i1 and i2.
Calculate the similarity between i1 and i2.
We will build this filtering system by splitting the dataset into 80% training set and 20% test set.

```{r}
sampled_data<- sample(x = c(TRUE, FALSE),
                      size = nrow(movie_ratings),
                      replace = TRUE,
                      prob = c(0.8, 0.2))
training_data <- movie_ratings[sampled_data, ]
testing_data <- movie_ratings[!sampled_data, ]
```

Building the Recommendation System using R
We will now explore the various parameters of our Item Based Collaborative Filter. These parameters are default in nature. In the first step, k denotes the number of items for computing their similarities. Here, k is equal to 30. Therefore, the algorithm will now identify the k most similar items and store their number. We use the cosine method which is the default one but you can also use pearson method.

```{r}
recommendation_system <- recommenderRegistry$get_entries(dataType ="realRatingMatrix")
recommendation_system$IBCF_realRatingMatrix$parameters

```

```{r}
recommen_model <- Recommender(data = training_data,
                          method = "IBCF",
                          parameter = list(k = 30))
recommen_model
class(recommen_model)
```
Let us now explore our data science recommendation system model as follows –

Using the getModel() function, we will retrieve the recommen_model. We will then find the class and dimensions of our similarity matrix that is contained within model_info. Finally, we will generate a heatmap, that will contain the top 20 items and visualize the similarity shared between them.

```{r}
model_info <- getModel(recommen_model)
class(model_info$sim)
dim(model_info$sim)
top_items <- 20
image(model_info$sim[1:top_items, 1:top_items],
   main = "Heatmap of the first rows and columns")

```

In the next step of ML project, we will carry out the sum of rows and columns with the similarity of the objects above 0. We will visualize the sum of columns through a distribution as follows –

```{r}
sum_rows <- rowSums(model_info$sim > 0)
table(sum_rows)

sum_cols <- colSums(model_info$sim > 0)
qplot(sum_cols, fill=I("steelblue"), col=I("red"))+ ggtitle("Distribution of the column count")
```


We will create a top_recommendations variable which will be initialized to 10, specifying the number of films to each user. We will then use the predict() function that will identify similar items and will rank them appropriately. Here, each rating is used as a weight. Each weight is multiplied with related similarities. Finally, everything is added in the end.

```{r}
top_recommendations <- 10 # the number of items to recommend to each user
predicted_recommendations <- predict(object = recommen_model,
                          newdata = testing_data,
                          n = top_recommendations)
predicted_recommendations
```
```{r}
user1 <- predicted_recommendations@items[[1]] # recommendation for the first user
movies_user1 <- predicted_recommendations@itemLabels[user1]
movies_user2 <- movies_user1
for (index in 1:10){
  movies_user2[index] <- as.character(subset(movie_data,
                                         movie_data$movieId == movies_user1[index])$title)
}
movies_user2
```
```{r}
recommendation_matrix <- sapply(predicted_recommendations@items,
                      function(x){ as.integer(colnames(movie_ratings)[x]) }) # matrix with the recommendations for each user
#dim(recc_matrix)
recommendation_matrix[,1:4]
```

